---
title: "Federal elections - 2022 and 2025"
author: "Peter GS"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include = F}

knitr::opts_chunk$set(echo = F)

```


```{r message = F}

library(tidyverse)
library(scales)  ## For number format on plot axes
library(kableExtra)

```



```{r}

## Load data

hvotes2022 <- read_csv("../data/House1stPrefsByCandidateByVoteType2022-27966-20250517.csv", skip = 1, show_col_types = FALSE)

hvotes2025 <- read_csv("../data/House1stPrefsByCandidateByVoteType2025-31496-20250517.csv", skip = 1, show_col_types = FALSE)

```

<br>

# House of Representatives

## Informal votes by state - 2022

```{r}

h_state_informal_22 <- 
  hvotes2022 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

h_state_informal_22 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

## Remove "message = F" to see what full_join is using to make the join.

h_state_votes_22 <- 
  hvotes2022 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_state_votes_22 <- 
  full_join(h_state_votes_22, h_state_informal_22) %>%
  #full_join(state_votes_22, state_informal_22), by = join_by(StateAb)) %>%
  mutate(state_formal = state_votes - state_informal)


N_hvotes_22 <- 
  h_state_votes_22 %>%
  summarise(sum(state_votes))

```

<br>

## Informal votes by state - 2025

```{r}

h_state_informal_25 <- 
  hvotes2025 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

h_state_informal_25 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

h_state_votes_25 <- 
  hvotes2025 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_state_votes_25 <- 
  full_join(h_state_votes_25, h_state_informal_25) %>%
  mutate(state_formal = state_votes - state_informal)

N_hvotes_25 <- 
  h_state_votes_25 %>%
  summarise(sum(state_votes))

```

<br>

## Difference in total number of ballot papers lodged between 2022 and 2025

```{r}

h_diff_N <- number(as.numeric(N_hvotes_25 - N_hvotes_22), big.mark = ",")


##diff_N

```

There were `r h_diff_N` more voters who voted for the House of 
Reps in 2025 than in 2022.

```{r}

party_nm <- 
  hvotes2022 %>%
  distinct(PartyNm)
  
#party_nm

```

<br>

## Votes by state - 2022 and 2025

```{r}

h_state_votes_2225 <-
  bind_rows(h_state_votes_22, h_state_votes_25) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal) %>%
  arrange(StateAb, Year)

h_state_votes_2225 %>%
  mutate(across(state_votes:state_formal, ~ number(.x, big.mark = ","))) %>%
  kable(booktabs = T, col.names = names(h_state_votes_2225), 
        align = "lcrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)


```
<br>

```{r}

ggplot(h_state_votes_2225, aes(x = StateAb, y = state_votes, 
                               shape = factor(Year))) +
  #geom_point() +
  geom_jitter(width = 0.133) +
  scale_y_continuous(labels = label_comma()) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: Total no. of ballot papers received (formal & informal)", 
       subtitle = "AEC data at 2025-05-18", 
       x = "State", y = "No. of votes") 
         
```

<br>

## House of Reps: Percentage of first preference votes for ALP - 2022 and 2025

```{r message = F}

h_alp_n_22 <-
  hvotes2022 %>%
  filter(PartyAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_alp_n_22 <- full_join(h_alp_n_22, h_state_votes_22)

## Note full_join uses: Joining with `by = join_by(StateAb, Year)` 

h_alp_n_25 <-
  hvotes2025 %>%
  filter(PartyAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_alp_n_25 <- full_join(h_alp_n_25, h_state_votes_25)

h_alp_n_2225 <-
  bind_rows(h_alp_n_22, h_alp_n_25)


```


```{r message = F}

h_alp_pc_2225 <- 
  full_join(h_alp_n_2225, h_state_votes_2225) %>%
  mutate(alp_pc = alp_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal,
         alp_votes, alp_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

h_alp_pc_2225 %>%
  mutate(across(state_votes:alp_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("alp_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(h_alp_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

## Display plot

ggplot(h_alp_pc_2225, aes(x = StateAb, y = alp_pc, 
                          shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: ALP - Percentage of first preference votes", 
       subtitle = "AEC data at 2025-05-18", 
       x = "State", y = "Percentage of votes")


## Save plot

ggsave(file = "../figs/house_alp_first_pref.pdf", dev = "pdf")

```

<br>

```{r}

## Check party names in 2022 and 2025

hvotes2022 %>%
  select(PartyAb:PartyNm) %>%
  distinct(PartyAb, PartyNm, .keep_all = T) %>%
  write_csv(., file = "party_names_22.csv")

hvotes2025 %>%
  select(PartyAb:PartyNm) %>%
  distinct(PartyAb, PartyNm, .keep_all = T) %>%
  write_csv(., file = "party_names_25.csv")

```


```{r include = F}

hvotes2022 %>%
  filter(PartyNm == "A.L.P.")

```

<br>

## House of Reps: Percentage of first preference votes for the Greens - 2022 and 2025

There are three separate figures for the Greens in the AEC voting figures: Australian Greens, Qld Greens and WA Greens. 

```{r message = F}

h_grn_n_22 <-
  hvotes2022 %>%
  filter(PartyAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_grn_n_22 <-
  full_join(h_grn_n_22, h_state_votes_22) %>%
  write_csv(., file = "h_grn_n_22.csv")


h_grn_n_25 <-
  hvotes2025 %>%
  filter(PartyAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_grn_n_25 <-
  full_join(h_grn_n_25, h_state_votes_25) %>%
  write_csv(., file = "h_grn_n_25.csv")


h_grn_n_2225 <-
  bind_rows(h_grn_n_22, h_grn_n_25)

```


```{r message = F}

h_grn_pc_2225 <- 
  full_join(h_grn_n_2225, h_state_votes_2225) %>%
  mutate(grn_pc = grn_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal,
         grn_votes, grn_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

h_grn_pc_2225 %>%
  mutate(across(state_votes:grn_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("grn_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(h_grn_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(h_grn_pc_2225, aes(x = StateAb, y = grn_pc, 
                          shape = factor(Year))) +
  geom_point() +
  scale_y_continuous(breaks = seq(10, 20, 2)) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: Greens - Percentage of first preference votes", 
       subtitle = "AEC data at 2025-05-18",
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/house_grn_first_pref.pdf", dev = "pdf")


```

<br>

## House of Reps NSW: Greens votes in 2022 and 2025

```{r}

grn_n_div_2022 <-
  hvotes2022 %>%
  filter(StateAb == "NSW") %>%
  filter(PartyAb == "GRN") %>%
  select(StateAb, PartyAb, DivisionNm, TotalVotes) %>%
  mutate(Year = "2022") %>%
  write_csv(., file = "grn_n_div_2022.csv")

```



```{r}

grn_n_div_2025 <-
  hvotes2025 %>%
  filter(StateAb == "NSW") %>%
  filter(PartyAb == "GRN") %>%
  select(StateAb, PartyAb, DivisionNm, TotalVotes) %>%
  mutate(Year = "2025") %>%
  write_csv(., file = "grn_n_div_2025.csv")

```


```{r eval = F, message = F}

ggplot(grn_n_div_2025, aes(x = DivisionNm, y = TotalVotes)) +
  geom_point() +
  labs(title = "House of Reps NSW: Greens votes - 2025",
       subtitle = "AEC data at 2025-05-18",
       x = "Electorate", y = "No. of votes") +
  theme(axis.text.x = element_text(angle = 75))

## Save plot

ggsave(file = "../figs/house_grn_first_pref.pdf", dev = "pdf")

```


```{r}

grn_n_div_2225 <- bind_rows(grn_n_div_2022, grn_n_div_2025) %>%
  write_csv(., file = "grn_n_div_2225.csv")

```


```{r out.width = "120%", message = F}

ggplot(grn_n_div_2225, aes(x = DivisionNm, y = TotalVotes,
                           shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps NSW: Greens first preference votes - 2022 and 2025",
       subtitle = "AEC data at 2025-05-18",
       x = "Electorate", y = "No. of votes") +
  theme(axis.text.x = element_text(angle = 75, vjust = 0.5))

## Save plot

ggsave(file = "../figs/house_grn_first_pref_electorates.pdf", dev = "pdf")

```


<br>
<br>

# Senate

```{r}

## Load data

svotes2022 <- read_csv("../data/Senate1stPrefsByStateByGroupByVoteType2022-27966-20250518.csv", skip = 1, show_col_types = FALSE)

svotes2025 <- read_csv("../data/Senate1stPrefsByStateByGroupByVoteType2025-31496-20250518.csv", skip = 1, show_col_types = FALSE)

```

<br>

## Informal votes by state - 2022

Informal Senate votes have already been removed by the Australian Electoral Commission in the data available for importing.

```{r eval = F}

s_state_informal_22 <- 
  svotes2022 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

s_state_informal_22 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

## Remove "message = F" to see what full_join is using to make the join.

s_state_votes_22 <- 
  svotes2022 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

#s_state_votes_22 <- 
#  full_join(s_state_votes_22, s_state_informal_22) %>%
  ##full_join(state_votes_22, state_informal_22), by = join_by(StateAb)) %>%
#  mutate(state_formal = state_votes - state_informal)
#  mutate(state_formal = state_votes)

s_state_votes_22$state_formal <- s_state_votes_22$state_votes
  
N_svotes_22 <- 
  s_state_votes_22 %>%
  summarise(sum(state_votes))

```

<br>

## Informal votes by state - 2025

Informal Senate votes have already been removed by the Australian Electoral Commission in the data available for importing.

```{r eval = F}

s_state_informal_25 <- 
  svotes2025 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

s_state_informal_25 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

s_state_votes_25 <- 
  svotes2025 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

#s_state_votes_25 <- 
#  full_join(s_state_votes_25, s_state_informal_25) %>%
#  mutate(state_formal = state_votes - state_informal)

s_state_votes_25$state_formal <- s_state_votes_25$state_votes

N_svotes_25 <- 
  s_state_votes_25 %>%
  summarise(sum(state_votes))

```

<br>

## Difference in total number of ballot papers lodged between 2022 and 2025

```{r}

s_diff_N <- number(as.numeric(N_svotes_25 - N_svotes_22), big.mark = ",")

##s_diff_N

```

There were `r s_diff_N` more people who lodged formal votes for the Senate in 2025 than in 2022.

```{r}

party_nm <- 
  svotes2022 %>%
  distinct(GroupNm)
  
#party_nm

```

<br>

## Votes by state - 2022 and 2025

```{r}

s_state_votes_2225 <-
  bind_rows(s_state_votes_22, s_state_votes_25) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal) %>%
  arrange(StateAb, Year)

s_state_votes_2225 %>%
  mutate(across(state_votes:state_formal, ~ number(.x, big.mark = ","))) %>%
  kable(booktabs = T, col.names = names(s_state_votes_2225), 
        align = "lcrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)


```

<br>

```{r}

ggplot(s_state_votes_2225, aes(x = StateAb, y = state_votes, 
                               shape = factor(Year))) +
  #geom_point() +
  geom_jitter(width = 0.133) +
  scale_y_continuous(labels = label_comma()) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: Total no. of formal votes", 
       subtitle = "AEC data at 2025-05-18",
       x = "State", y = "No. of formal votes")
         
```

<br>

## Senate: Percentage of first preference votes for ALP - 2022 and 2025

```{r message = F}

s_alp_n_22 <-
  svotes2022 %>%
  filter(GroupAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

s_alp_n_22 <- full_join(s_alp_n_22, s_state_votes_22)

## Note full_join uses: Joining with `by = join_by(StateAb, Year)` 

s_alp_n_25 <-
  svotes2025 %>%
  filter(GroupAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

s_alp_n_25 <- full_join(s_alp_n_25, s_state_votes_25)

s_alp_n_2225 <-
  bind_rows(s_alp_n_22, s_alp_n_25)


```


```{r message = F}

s_alp_pc_2225 <- 
  full_join(s_alp_n_2225, s_state_votes_2225) %>%
  mutate(alp_pc = alp_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal,
         alp_votes, alp_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

s_alp_pc_2225 %>%
  mutate(across(state_votes:alp_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("alp_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(s_alp_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(s_alp_pc_2225, aes(x = StateAb, y = alp_pc, 
                          shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: ALP - Percentage of first preference votes", 
       subtitle = "AEC data at 2025-05-18",
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/senate_alp_first_pref.pdf", dev = "pdf")

```

<br>

```{r}

## Check party names in 2022 and 2025

svotes2022 %>%
  select(GroupAb:GroupNm) %>%
  distinct(GroupAb, GroupNm, .keep_all = T) %>%
  write_csv(., file = "party_names_22.csv")

svotes2025 %>%
  select(GroupAb:GroupNm) %>%
  distinct(GroupAb, GroupNm, .keep_all = T) %>%
  write_csv(., file = "party_names_25.csv")

```


```{r include = F}

svotes2022 %>%
  filter(GroupNm == "A.L.P.")

```

<br>

## Senate: Percentage of first preference votes for the Greens - 2022 and 2025

The Greens go by different party names, depending on their state: Australian Greens, Qld Greens and WA Greens. All are identified by the same abbreviation, GRN, in the AEC data.

```{r message = F}

s_grn_n_22 <-
  svotes2022 %>%
  filter(GroupAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

s_grn_n_22 <-
  full_join(s_grn_n_22, s_state_votes_22) %>%
  write_csv(., file = "s_grn_n_22.csv")


s_grn_n_25 <-
  svotes2025 %>%
  filter(GroupAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

s_grn_n_25 <-
  full_join(s_grn_n_25, s_state_votes_25) %>%
  write_csv(., file = "s_grn_n_25.csv")


s_grn_n_2225 <-
  bind_rows(s_grn_n_22, s_grn_n_25)

```



```{r message = F}

s_grn_pc_2225 <- 
  full_join(s_grn_n_2225, s_state_votes_2225) %>%
  mutate(grn_pc = grn_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal,
         grn_votes, grn_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

s_grn_pc_2225 %>%
  mutate(across(state_votes:grn_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("grn_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(s_grn_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(s_grn_pc_2225, aes(x = StateAb, y = grn_pc, 
                          shape = factor(Year))) +
  geom_point() +
  scale_y_continuous(breaks = seq(10, 20, 2)) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: Greens - Percentage of first preference votes", 
       subtitle = "AEC data at 2025-05-18", 
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/senate_grn_first_pref.pdf", dev = "pdf")

```


