---
title: "Federal elections - 2022 and 2025"
author: "Peter GS"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
    toc: true
    toc_depth: 3
    toc_float: true
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include = F}

knitr::opts_chunk$set(echo = F)

```


```{r message = F}

library(tidyverse)
library(scales)  ## For number format on plot axes
library(kableExtra)
library(ggrepel)  ## For labelling data point

```



```{r eval = T}

## Download data directly from AEC

hvotes2022_web <- read_csv("https://results.aec.gov.au/27966/Website/Downloads/HouseFirstPrefsByCandidateByVoteTypeDownload-27966.csv", 
                           skip = 1, show_col_types = FALSE)

## Save data file

f_name1 <- paste0("../data/hvotes2022_web_", Sys.Date(), ".csv")

write_csv(hvotes2022_web, file = f_name1)



hvotes2025_web <- read_csv("https://tallyroom.aec.gov.au/Downloads/HouseFirstPrefsByCandidateByVoteTypeDownload-31496.csv", 
                           skip = 1, show_col_types = FALSE)

## Save data file

f_name2 <- paste0("../data/hvotes2025_web_", Sys.Date(), ".csv")

write_csv(hvotes2025_web, file = f_name2)


```



```{r eval = F}

## Load local data

hvotes2022_local <- read_csv("../data/hvotes2022_web_20250520.csv", skip = 0)

hvotes2025_local <- read_csv("../data/hvotes2025_web_20250520.csv", skip = 0)

```


```{r}

## Pick the data set to use!!

hvotes2022 <- hvotes2022_web

#hvotes2022 <- hvotes2022_local


hvotes2025 <- hvotes2025_web

#hvotes2025 <- hvotes2025_local

```


```{r}

## Set date of data for titles in plots

date_data <- Sys.Date()

```

<br>

# House of Representatives

## Informal votes by state - 2022

```{r}

h_state_informal_22 <- 
  hvotes2022 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

h_state_informal_22 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

## Remove "message = F" to see what full_join is using to make the join.

h_state_votes_22 <- 
  hvotes2022 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_state_votes_22 <- 
  full_join(h_state_votes_22, h_state_informal_22) %>%
  #full_join(state_votes_22, state_informal_22), by = join_by(StateAb)) %>%
  mutate(state_formal = state_votes - state_informal)


N_hvotes_22 <- 
  h_state_votes_22 %>%
  summarise(sum(state_votes))

```

<br>

## Informal votes by state - 2025

```{r}

h_state_informal_25 <- 
  hvotes2025 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

h_state_informal_25 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

h_state_votes_25 <- 
  hvotes2025 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_state_votes_25 <- 
  full_join(h_state_votes_25, h_state_informal_25) %>%
  mutate(state_formal = state_votes - state_informal)

N_hvotes_25 <- 
  h_state_votes_25 %>%
  summarise(sum(state_votes))

```

<br>

## Difference in total number of ballot papers lodged between 2022 and 2025

```{r}

h_diff_N <- number(as.numeric(N_hvotes_25 - N_hvotes_22), big.mark = ",")


##diff_N

```

There were `r h_diff_N` more voters who voted for the House of 
Reps in 2025 than in 2022.

```{r}

party_nm <- 
  hvotes2022 %>%
  distinct(PartyNm)
  
#party_nm

```

<br>

## Votes by state - 2022 and 2025

```{r}

h_state_votes_2225 <-
  bind_rows(h_state_votes_22, h_state_votes_25) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal) %>%
  arrange(StateAb, Year)

h_state_votes_2225 %>%
  mutate(across(state_votes:state_formal, ~ number(.x, big.mark = ","))) %>%
  kable(booktabs = T, col.names = names(h_state_votes_2225), 
        align = "lcrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)


```
<br>



```{r}

ggplot(h_state_votes_2225, aes(x = StateAb, y = state_votes, 
                               shape = factor(Year))) +
  #geom_point() +
  geom_jitter(width = 0.133) +
  scale_y_continuous(labels = label_comma()) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: Total no. of ballot papers received (formal & informal)", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "),
       x = "State", y = "No. of votes") 
         
```

<br>

## House of Reps: Percentage of first preference votes for ALP - 2022 and 2025

```{r message = F}

h_alp_n_22 <-
  hvotes2022 %>%
  filter(PartyAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_alp_n_22 <- full_join(h_alp_n_22, h_state_votes_22)

## Note full_join uses: Joining with `by = join_by(StateAb, Year)` 

h_alp_n_25 <-
  hvotes2025 %>%
  filter(PartyAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_alp_n_25 <- full_join(h_alp_n_25, h_state_votes_25)

h_alp_n_2225 <-
  bind_rows(h_alp_n_22, h_alp_n_25)


```


```{r message = F}

h_alp_pc_2225 <- 
  full_join(h_alp_n_2225, h_state_votes_2225) %>%
  mutate(alp_pc = alp_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal,
         alp_votes, alp_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

h_alp_pc_2225 %>%
  mutate(across(state_votes:alp_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("alp_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(h_alp_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

## Display plot

ggplot(h_alp_pc_2225, aes(x = StateAb, y = alp_pc, 
                          shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: ALP - Percentage of first preference votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "Percentage of votes")


## Save plot

ggsave(file = "../figs/house_alp_first_pref.pdf", dev = "pdf")

```

<br>

```{r}

## Check party names in 2022 and 2025

hvotes2022 %>%
  select(PartyAb:PartyNm) %>%
  distinct(PartyAb, PartyNm, .keep_all = T) %>%
  write_csv(., file = "party_names_22.csv")

hvotes2025 %>%
  select(PartyAb:PartyNm) %>%
  distinct(PartyAb, PartyNm, .keep_all = T) %>%
  write_csv(., file = "party_names_25.csv")

```


```{r include = F}

hvotes2022 %>%
  filter(PartyNm == "A.L.P.")

```

<br>

## House of Reps: Percentage of first preference votes for the Greens - 2022 and 2025

There are three separate figures for the Greens in the AEC voting figures: Australian Greens, Qld Greens and WA Greens. 

```{r message = F}

h_grn_n_22 <-
  hvotes2022 %>%
  filter(PartyAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

h_grn_n_22 <-
  full_join(h_grn_n_22, h_state_votes_22) %>%
  write_csv(., file = "h_grn_n_22.csv")


h_grn_n_25 <-
  hvotes2025 %>%
  filter(PartyAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

h_grn_n_25 <-
  full_join(h_grn_n_25, h_state_votes_25) %>%
  write_csv(., file = "h_grn_n_25.csv")


h_grn_n_2225 <-
  bind_rows(h_grn_n_22, h_grn_n_25)

```


```{r message = F}

h_grn_pc_2225 <- 
  full_join(h_grn_n_2225, h_state_votes_2225) %>%
  mutate(grn_pc = grn_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes, state_informal, state_formal,
         grn_votes, grn_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

h_grn_pc_2225 %>%
  mutate(across(state_votes:grn_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("grn_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(h_grn_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(h_grn_pc_2225, aes(x = StateAb, y = grn_pc, 
                          shape = factor(Year))) +
  geom_point() +
  scale_y_continuous(breaks = seq(10, 20, 2)) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps: Greens - Percentage of first preference votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/house_grn_first_pref.pdf", dev = "pdf")


```

<br>

## House of Reps NSW: Greens votes in 2022 and 2025


```{r}

## Green votes in NSW - 2022

N_h_nsw_div_2022 <-
  hvotes2022 %>%
  filter(StateAb == "NSW") %>%
  filter(!Surname == "Informal") %>%  ## Don't use PartyNm!
  filter(!DivisionNm == "North Sydney") %>%group_by(DivisionNm) %>%
  summarise(TotalVotesDIV = sum(TotalVotes))


grn_h_n_nsw_div_2022a <-
  hvotes2022 %>%
  filter(StateAb == "NSW") %>%
  filter(PartyAb == "GRN") %>%
  select(StateAb, PartyAb, DivisionNm, TotalVotes) %>%
  rename(TotalVotesGRN = TotalVotes) %>%
  mutate(Year = "2022") 


grn_h_n_nsw_div_2022 <-
  inner_join(N_h_nsw_div_2022, grn_h_n_nsw_div_2022a, by = "DivisionNm") %>%
  mutate(grn_pc = TotalVotesGRN / TotalVotesDIV * 100) %>%
  select(StateAb, Year, PartyAb, DivisionNm, TotalVotesGRN, TotalVotesDIV,
         grn_pc) %>%
  write_csv(., file = "grn_h_n_nsw_div_2022.csv")


```


```{r}

## Green votes in NSW - 2025

N_h_nsw_div_2025 <-
  hvotes2025 %>%
  filter(StateAb == "NSW") %>%
  filter(!Surname == "Informal") %>% ## Don't use PartyNm!
  group_by(DivisionNm) %>%
  summarise(TotalVotesDIV = sum(TotalVotes))


grn_h_n_nsw_div_2025a <-
  hvotes2025 %>%
  filter(StateAb == "NSW") %>%
  filter(PartyAb == "GRN") %>%
  select(StateAb, PartyAb, DivisionNm, TotalVotes) %>%
  rename(TotalVotesGRN = TotalVotes) %>%
  mutate(Year = "2025") 


grn_h_n_nsw_div_2025 <-
  inner_join(N_h_nsw_div_2025, grn_h_n_nsw_div_2025a, by = "DivisionNm") %>%
  mutate(grn_pc = TotalVotesGRN / TotalVotesDIV * 100) %>%
  select(StateAb, Year, PartyAb, DivisionNm, TotalVotesGRN, TotalVotesDIV,
         grn_pc) %>%
  write_csv(., file = "grn_h_n_nsw_div_2025.csv")

```


```{r message = F}

## Consolidated Greens votes in NSW - 2022 and 2025

grn_h_n_nsw_div_2225 <-
  full_join(grn_h_n_nsw_div_2022, grn_h_n_nsw_div_2025) %>%
  arrange(DivisionNm, Year) %>%
  write_csv(., file = "grn_h_n_nsw_div_2225.csv")


## Format data to display in a nice table

grn_h_n_nsw_div_2225 %>%
  mutate(across(TotalVotesGRN:TotalVotesDIV, ~ number(.x, big.mark = ","))) %>%
  mutate(across("grn_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(grn_h_n_nsw_div_2225), 
        align = "lcclrrr") %>%
  #row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  row_spec(c(seq(3, 91, 4), seq(4, 92, 4)), background = "#eeeeee") %>%
  kable_styling(full_width = F)


```

<br>

```{r out.width = "120%", message = F}

ggplot(grn_h_n_nsw_div_2225, aes(x = DivisionNm, y = TotalVotesGRN, 
                                 shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps NSW: Greens first preference votes - 2022 and 2025",
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "Electorate", y = "No. of votes") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

## Save plot

ggsave(file = "../figs/house_grn_first_pref_electorates.pdf", dev = "pdf")

```

<br>


```{r out.width = "120%", message = F}

ggplot(grn_h_n_nsw_div_2225, aes(x = DivisionNm, y = grn_pc, 
                                 shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "House of Reps NSW: Greens percent. first preference votes - 2022 and 2025",
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "Electorate", y = "Percentage of votes in electorate") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

## Save plot

ggsave(file = "../figs/house_grn_first_pref_pc_electorates.pdf", dev = "pdf")

```

<br>

## House of Reps NSW: First preference by polling place

```{r}

hvotes2025_poll_place <- read_csv("https://tallyroom.aec.gov.au/Downloads/HouseStateFirstPrefsByPollingPlaceDownload-31496-NSW.csv", skip = 1, show_col_types = F)

## Save data file

f_name5 <- paste0("../data/hvotes2025_poll_place_", Sys.Date(), ".csv")

write_csv(hvotes2025_poll_place, file = f_name5)

  
```


```{r}

hvotes2025_elect_macq_grn <-
  hvotes2025_poll_place %>%
  filter(Surname != "Informal") %>%
  filter(DivisionNm == "Macquarie") %>%
  group_by(PollingPlace) %>%
  mutate(TotalPollPlaceVotes = sum(OrdinaryVotes)) %>%
  mutate(party_pc = OrdinaryVotes / sum(OrdinaryVotes) * 100)  %>% 
  filter(PartyAb == "GRN")

write_csv(hvotes2025_elect_macq_grn, file = "hvotes2025_macq_grn.csv")

```


```{r}

## Shorten long polling place names

hvotes2025_elect_macq_grn <-
  hvotes2025_elect_macq_grn %>%
  mutate(PollingPlace2 = str_replace(PollingPlace, "Macquarie", "")) %>%
  mutate(PollingPlace2 = str_replace(PollingPlace2, "MACQUARIE", "")) %>%
  mutate(PollingPlace2 = str_replace(PollingPlace2, "\\(\\)", "")) %>%
  mutate(PollingPlace2 = str_replace(PollingPlace2, 
                                     "Special Hospital Team 1",
                                     "Spec. Hospital 1")) %>% 
  mutate(PollingPlace2 = str_replace(PollingPlace2, 
                                     "Special Hospital Team 2",
                                     "Spec. Hospital 2"))


```


```{r out.width = "120%", message = F}

#hvotes2025_elect_grn_pc$PollingPlace[1:31]

ggplot(hvotes2025_elect_macq_grn, aes(x = PollingPlace2, y = party_pc)) +
  #geom_point(aes(colour = TotalPollPlaceVotes < 200)) +
  geom_point() +
  #geom_text(aes(label = OrdinaryVotes), size = 2, nudge_y = -1.25) +
  ggrepel::geom_text_repel(aes(label = OrdinaryVotes), size = 2,
                           box.padding = 0.1) +
  labs(title = "House of Reps Macquarie: Greens percent. first prefs. by polling place - 2025",
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("Vote numbers shown near data point. ", "AEC data at", date_data, sep = " "), 
       x = "Polling Place", y = "Percentage of votes at polling place") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

## Save plot

ggsave(file = "../figs/house_grn_first_pref_pc_macquarie.pdf", dev = "pdf")


```


```{r eval = F}

hvotes2025_elect_macq_grn %>%
  filter(grepl("Springwood", PollingPlace)) %>%
  select(PollingPlace, OrdinaryVotes)

```

<br>
<br>

# Senate

```{r}

## Download data directly from AEC

svotes2022_web <- read_csv("https://results.aec.gov.au/27966/Website/Downloads/SenateFirstPrefsByStateByGroupByVoteTypeDownload-27966.csv", 
                           skip = 1, show_col_types = FALSE)

## Save data file

f_name3 <- paste0("../data/svotes2022_web_", Sys.Date(), ".csv")

write_csv(svotes2022_web, file = f_name3)



svotes2025_web <- read_csv("https://tallyroom.aec.gov.au/Downloads/SenateFirstPrefsByStateByGroupByVoteTypeDownload-31496.csv", 
                           skip = 1, show_col_types = FALSE)

## Save data file

f_name4 <- paste0("../data/svotes2025_web_", Sys.Date(), ".csv")

write_csv(svotes2025_web, file = f_name4)


```



```{r eval = F}

## Load local data

svotes2022_local <- read_csv("../data/svotes2022_web_20250520.csv", skip = 0)

svotes2025_local <- read_csv("../data/svotes2025_web_20250520.csv", skip = 0)

```


```{r}

## Pick the data set to use!!

svotes2022 <- svotes2022_web

#svotes2022 <- svotes2022_local


svotes2025 <- svotes2025_web

#svotes2025 <- svotes2025_local


```


<br>

## Informal votes by state - 2022

Informal Senate votes have already been removed by the Australian Electoral Commission in the data available for importing.

```{r eval = F}

s_state_informal_22 <- 
  svotes2022 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

s_state_informal_22 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

## Remove "message = F" to see what full_join is using to make the join.

s_state_votes_22 <- 
  svotes2022 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

#s_state_votes_22 <- 
#  full_join(s_state_votes_22, s_state_informal_22) %>%
  ##full_join(state_votes_22, state_informal_22), by = join_by(StateAb)) %>%
#  mutate(state_formal = state_votes - state_informal)
#  mutate(state_formal = state_votes)

s_state_votes_22$state_formal <- s_state_votes_22$state_votes
  
N_svotes_22 <- 
  s_state_votes_22 %>%
  summarise(sum(state_votes))

```

<br>

## Informal votes by state - 2025

Informal Senate votes have already been removed by the Australian Electoral Commission in the data available for importing.

```{r eval = F}

s_state_informal_25 <- 
  svotes2025 %>%
  filter(Surname == "Informal") %>%
  group_by(StateAb) %>%
  summarise(state_informal = sum(TotalVotes))

s_state_informal_25 %>%
  kbl() %>%
  kable_styling(full_width = F)

```


```{r message = F}

s_state_votes_25 <- 
  svotes2025 %>%
  group_by(StateAb) %>%
  summarise(state_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

#s_state_votes_25 <- 
#  full_join(s_state_votes_25, s_state_informal_25) %>%
#  mutate(state_formal = state_votes - state_informal)

s_state_votes_25$state_formal <- s_state_votes_25$state_votes

N_svotes_25 <- 
  s_state_votes_25 %>%
  summarise(sum(state_votes))

```

<br>

## Difference in total number of ballot papers lodged between 2022 and 2025

```{r}

s_diff_N <- number(as.numeric(N_svotes_25 - N_svotes_22), big.mark = ",")

##s_diff_N

```

There were `r s_diff_N` more people who lodged formal votes for the Senate in 2025 than in 2022.

```{r}

party_nm <- 
  svotes2022 %>%
  distinct(GroupNm)
  
#party_nm

```

<br>

## Votes by state - 2022 and 2025

```{r}

s_state_votes_2225 <-
  bind_rows(s_state_votes_22, s_state_votes_25) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal) %>%
  arrange(StateAb, Year)

s_state_votes_2225 %>%
  mutate(across(state_votes:state_formal, ~ number(.x, big.mark = ","))) %>%
  kable(booktabs = T, col.names = names(s_state_votes_2225), 
        align = "lcrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)


```

<br>

```{r}

ggplot(s_state_votes_2225, aes(x = StateAb, y = state_votes, 
                               shape = factor(Year))) +
  #geom_point() +
  geom_jitter(width = 0.133) +
  scale_y_continuous(labels = label_comma()) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: Total no. of formal votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "No. of formal votes")
         
```

<br>

## Senate: Percentage of first preference votes for ALP - 2022 and 2025

```{r message = F}

s_alp_n_22 <-
  svotes2022 %>%
  filter(GroupAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

s_alp_n_22 <- full_join(s_alp_n_22, s_state_votes_22)

## Note full_join uses: Joining with `by = join_by(StateAb, Year)` 

s_alp_n_25 <-
  svotes2025 %>%
  filter(GroupAb == "ALP") %>%
  group_by(StateAb) %>%
  summarise(alp_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

s_alp_n_25 <- full_join(s_alp_n_25, s_state_votes_25)

s_alp_n_2225 <-
  bind_rows(s_alp_n_22, s_alp_n_25)


```


```{r message = F}

s_alp_pc_2225 <- 
  full_join(s_alp_n_2225, s_state_votes_2225) %>%
  mutate(alp_pc = alp_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal,
         alp_votes, alp_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

s_alp_pc_2225 %>%
  mutate(across(state_votes:alp_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("alp_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(s_alp_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(s_alp_pc_2225, aes(x = StateAb, y = alp_pc, 
                          shape = factor(Year))) +
  geom_point() +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: ALP - Percentage of first preference votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/senate_alp_first_pref.pdf", dev = "pdf")

```

<br>

```{r}

## Check party names in 2022 and 2025

svotes2022 %>%
  select(GroupAb:GroupNm) %>%
  distinct(GroupAb, GroupNm, .keep_all = T) %>%
  write_csv(., file = "party_names_22.csv")

svotes2025 %>%
  select(GroupAb:GroupNm) %>%
  distinct(GroupAb, GroupNm, .keep_all = T) %>%
  write_csv(., file = "party_names_25.csv")

```


```{r include = F}

svotes2022 %>%
  filter(GroupNm == "A.L.P.")

```

<br>

## Senate: Percentage of first preference votes for the Greens - 2022 and 2025

The Greens go by different party names, depending on their state: Australian Greens, Qld Greens and WA Greens. All are identified by the same abbreviation, GRN, in the AEC data.

```{r message = F}

s_grn_n_22 <-
  svotes2022 %>%
  filter(GroupAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

s_grn_n_22 <-
  full_join(s_grn_n_22, s_state_votes_22) %>%
  write_csv(., file = "s_grn_n_22.csv")


s_grn_n_25 <-
  svotes2025 %>%
  filter(GroupAb == "GRN") %>%
  group_by(StateAb) %>%
  summarise(grn_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

s_grn_n_25 <-
  full_join(s_grn_n_25, s_state_votes_25) %>%
  write_csv(., file = "s_grn_n_25.csv")


s_grn_n_2225 <-
  bind_rows(s_grn_n_22, s_grn_n_25)

```



```{r message = F}

s_grn_pc_2225 <- 
  full_join(s_grn_n_2225, s_state_votes_2225) %>%
  mutate(grn_pc = grn_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal,
         grn_votes, grn_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

s_grn_pc_2225 %>%
  mutate(across(state_votes:grn_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("grn_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(s_grn_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F}

ggplot(s_grn_pc_2225, aes(x = StateAb, y = grn_pc, 
                          shape = factor(Year))) +
  geom_point() +
  scale_y_continuous(breaks = seq(10, 20, 2)) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: Greens - Percentage of first preference votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/senate_grn_first_pref.pdf", dev = "pdf")

```

<br>


## Senate: Percentage of first preference votes for One Nation - 2022 and 2025


```{r message = F}

s_on_n_22 <-
  svotes2022 %>%
  filter(GroupAb == "ON") %>%
  group_by(StateAb) %>%
  summarise(on_votes = sum(TotalVotes)) %>%
  mutate(Year = "2022")

s_on_n_22 <-
  full_join(s_on_n_22, s_state_votes_22) %>%
  write_csv(., file = "s_on_n_22.csv")


s_on_n_25 <-
  svotes2025 %>%
  filter(GroupAb == "ON") %>%
  group_by(StateAb) %>%
  summarise(on_votes = sum(TotalVotes)) %>%
  mutate(Year = "2025")

s_on_n_25 <-
  full_join(s_on_n_25, s_state_votes_25) %>%
  write_csv(., file = "s_on_n_25.csv")


s_on_n_2225 <-
  bind_rows(s_on_n_22, s_on_n_25)

```



```{r message = F}

s_on_pc_2225 <- 
  full_join(s_on_n_2225, s_state_votes_2225) %>%
  mutate(on_pc = on_votes / state_formal * 100) %>%
  select(StateAb, Year, state_votes,
         #state_informal,
         state_formal,
         on_votes, on_pc) %>%
  arrange(StateAb, Year)

## "select" must go into code above so columns are in right order 
##   for section below

## Format the data to display a nice table

s_on_pc_2225 %>%
  mutate(across(state_votes:on_votes, ~ number(.x, big.mark = ","))) %>%
  mutate(across("on_pc", ~ round(.x, 1))) %>%
  kable(booktabs = T, col.names = names(s_on_pc_2225), 
        align = "lcrrrrr") %>%
  row_spec(c(3, 4, 7, 8, 11, 12, 15, 16), background = "#eeeeee") %>%
  kable_styling(full_width = F)
         

```

<br>


```{r message = F, warning = F}

ggplot(s_on_pc_2225, aes(x = StateAb, y = on_pc, 
                          shape = factor(Year))) +
  geom_point() +
  #scale_y_continuous(breaks = seq(10, 20, 2)) +
  guides(shape = guide_legend(title = "Year")) +
  labs(title = "Senate: One Nation - Percentage of first preference votes", 
       #subtitle = "AEC data at 2025-05-18",
       subtitle = paste("AEC data at", date_data, sep = " "), 
       x = "State", y = "Percentage of votes")

## Save plot

ggsave(file = "../figs/senate_on_first_pref.pdf", dev = "pdf")

```


